---
title: "data exploration"
author: "Marie Moriarty"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Import MAGeCK data set

```{r}
# Import MAGeCK data set
library(readr)
mageck <- read_delim("mageckRRA.gene_summary.txt", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)

# Reassign column names
colnames(mageck) <- c("id",
                      "num",
                      "neg.score",
                      "neg.p_value",
                      "neg.fdr",
                      "neg.rank",
                      "neg.goodsgrna",
                      "neg.lfc",
                      "pos.score",
                      "pos.p_value",
                      "pos.fdr",
                      "pos.rank",
                      "pos.goodsgrna",
                      "pos.lfc"
                      )

# Convert goodsgrna to factor
mageck$neg.goodsgrna <- as.factor(mageck$neg.goodsgrna)
mageck$pos.goodsgrna <- as.factor(mageck$pos.goodsgrna)


# view data summary
summary(mageck)
```

#### Initial plots
```{r}
library(ggplot2)

# Compare positive and negative group scores
ggplot(mageck, aes(x = neg.score, 
                   y = pos.score, 
                   alpha = 0.05)) +
  geom_point()

ggplot(mageck, aes(x = neg.score)) +
  geom_histogram(bins = 500)

ggplot(mageck, aes(x = pos.score)) +
  geom_histogram(bins = 500)

cor(mageck$neg.score, mageck$pos.score)

# Compare positive and negative p-values
ggplot(mageck, aes(x = neg.p_value, 
                   y = pos.p_value, 
                   alpha = 0.05)) +
  geom_point()

ggplot(mageck, aes(x = neg.p_value)) +
  geom_histogram(bins = 500)

ggplot(mageck, aes(x = pos.p_value)) +
  geom_histogram(bins = 500)

cor(mageck$neg.p_value, mageck$pos.p_value)


histogram <- function(var){
  ggplot(mageck, aes(x = var)) +
    geom_histogram(bins = 500)
}

# Positive histograms
histogram(mageck$pos.score)
histogram(mageck$pos.p_value)
histogram(mageck$pos.fdr)
histogram(mageck$pos.lfc)

# Negative histograms
histogram(mageck$neg.score)
histogram(mageck$neg.p_value)
histogram(mageck$neg.fdr)
histogram(mageck$neg.lfc)
```

### Separate more significant negative and positive p-values

I'm trying to split the data set into two based on whether the positive or negative p-value is more significant (smaller). I know I can use a loop to go element by element, but I'm wondering if there is a more efficient way. 
```{r}
# pos.p = NULL
# neg.p = NULL
# # equal.p = NULL
# 
# sort_fun <- function(x){
#   if (x$pos.p_value < x$neg.p_value){
#     pos.p <- c(pos.p, x$pos.p_value)
#   }
#   else{
#     neg.p <- c(neg.p, x$neg.p_value)
#   }
# }
# 
# apply(X = mageck,
#       MARGIN = ,
#       FUN = sort_fun)

```


#### Pathways
```{r}
# Code adapted from https://stackoverflow.com/questions/6602881/text-file-to-list-in-r

# Read in pathways data as list and split elements into strings
gmt <- scan("m2.cp.v2022.1.Mm.symbols.gmt", what = "", sep = "\n")
pathways <- strsplit(gmt, "[[:space:]]+")

# Assign first entry to names of each list element
names(pathways) <- sapply(pathways, `[[`, 1)

# save urls to separate list for reference
source <- sapply(pathways, `[[`, 2)

# Remove two beginning reference rows
pathways <- lapply(pathways, `[`, -c(1:2))


# Alternative set using GSEABase
library(GSEABase)
gmt2 <- getGmt("m2.cp.v2022.1.Mm.symbols.gmt")
```



How to separate the mageck data frame according to pathways? Do I try to new objects combining the pathways and the mageck results or do I find a way to compare the two sets.